Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation

### 2022-03-29 intro.tex
added more refs and weak form in bembem

The PBE has been solved numerically with finite difference\cite{BakerETal2001,GilsonETal1985,JurrusETal2018,LiETal2019}, finite element\cite{HolstETal2012,BondEtal2010}, boundary element\cite{AltmanBardhanWhiteTidor09,LuETal2006,GengKrasny2013,CooperBardhanBarba2014}, and analytical\cite{YapHeadgordon2010,FelbergETal2017} methods.
In particular, the boundary element method (BEM) has proven to be very efficient for high accuracy calculations [GengKrasny2013,CooperBardhanBarba2014], mainly due to the precise description of the molecular surface and point charges. 
However, BEM is limited to constant material properties in each region, and the linear version of the PBE. 
Even though these limitations are acceptable in a wide range of applications, there are cases when BEM falls short, for example, if a variable permittivity is required inside the solute [cite], or the solute is highly charged such that the linear approximation breaks [cite?].

This article presents a methodology to overcome some of those limitations, through coupling boundary and finite elements (FEM-BEM).
This approach brings the best of both worlds: the flexibility of FEM and efficiency of BEM, all in an accurate description of the dissolved molecule.

>>>>

The PBE has been solved numerically with finite difference\cite{BakerETal2001,GilsonETal1985,JurrusETal2018,LiETal2019}, finite element\cite{HolstETal2012,BondEtal2010}, boundary element\cite{AltmanBardhanWhiteTidor09,LuETal2006,GengKrasny2013,CooperBardhanBarba2014}, and analytical\cite{YapHeadgordon2010,FelbergETal2017} methods.
In particular, the boundary element method (BEM) has proven to be very efficient for high accuracy calculations \cite{GengKrasny2013,CooperBardhanBarba2014}, mainly due to the precise description of the molecular surface and point charges. 
However, BEM is limited to constant material properties in each region, and the linear version of the PBE. 
Even though these limitations are acceptable in a wide range of applications, there are cases when BEM falls short, for example, if a variable permittivity is required inside the solute \cite{grant2001smooth,li2013dielectric}, or the solute is highly charged such that the linear approximation breaks\cite{FogolariETal1999}.

This article presents a methodology to overcome some of those limitations, by coupling finite and boundary elements methods (FEM-BEM coupling).
This approach brings the best of both worlds: the flexibility of FEM and efficiency of BEM, all in an accurate description of the dissolved molecule.

### 2022-03-29 methods.tex
added more refs and weak form in bembem

There are several possible boundary integral formulations of Equation \eqref{eq:pbe}.~\cite{search2022towards} The simplest form was presented by Yoon ºand Lenhoff in 1991,~\cite{YoonLenhoff1990} known as the \emph{direct} formulation, however, the resulting system is usually ill-conditioned. Here, we use the better conditioned formulation presented by Lu and co-workers.~\cite{LuETal2006,LuETal2009,debuhr2016dashmm}
Applying Green's second identity to Equation \eqref{eq:pbe} we get 
%
\begin{align} \label{eq:volume_potential}
\phi_{1}(\mathbf{r}_{\Omega_1})+ K_{L}^{\Omega_1}(\phi_{1,\Gamma}) -  V_{L}^{\Omega_1} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right) & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Omega_1} - \mathbf{r}_k|}  \quad \text{on $\Omega_1$,} \nonumber \\
\phi_{2}(\mathbf{r}_{\Omega_2}) - K_{Y}^{\Omega_2}(\phi_{2,\Gamma}) + V_{Y}^{\Omega_2} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) & = 0 \quad \text{on $\Omega_2$,}
\end{align}
%
where $\mathbf{r}_\Omega$ is evaluated anywhere in the correspoding domain, except $\Gamma$, and $\phi_{1,\Gamma} = \phi_1(\mathbf{r}_\Gamma)$ and $\phi_{2,\Gamma} = \phi_2(\mathbf{r}_\Gamma)$ are the potentials on $\Gamma$ as we approach from $\Omega_1$ and $\Omega_2$, respectively. $K$ and $V$ are the double- and single-layer potentials for the Laplace (subscript $L$) and Yukawa (subscript $Y$, also known as modified Helmholtz) kernels
%
\begin{align}\label{eq:single_double}
V^\Omega_{L,Y}(\varphi) = \oint_\Gamma g_{L,Y}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
K^\Omega_{L,Y}(\varphi) = \oint_\Gamma \frac{\partial g_{L,Y}}{\partial\mathbf{n}'}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}',\nonumber\\
\end{align}
%
where $\varphi(\mathbf{r})$ can be any distribution over $\Gamma$, and 
%
\begin{align}\label{eq:green_func}
g_L(\mathbf{r},\mathbf{r}')=\frac{1}{4\pi|\mathbf{r}-\mathbf{r}'|} \nonumber \\
g_Y(\mathbf{r},\mathbf{r}')=\frac{e^{-\kappa|\mathbf{r}-\mathbf{r}'|}}{4\pi|\mathbf{r}-\mathbf{r}'|}
\end{align}
%
are the free-space Green's function of the Laplace and linearized Poisson-Boltzmann equations, respectively.

If we take the limit of Eq. \eqref{eq:volume_potential} as $\mathbf{r}_{\Omega}$ approaches $\Gamma$, we get
%
\begin{align} \label{eq:direct}
\frac{\phi_{1,\Gamma}}{2}+ K_{L}^{\Gamma}(\phi_{1,\Gamma}) -  V_{L}^{\Gamma} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right) & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Gamma} - \mathbf{r}_k|} \nonumber \\
\frac{\phi_{2,\Gamma}}{2} - K_{Y}^{\Gamma}(\phi_{2,\Gamma}) + V_{Y}^{\Gamma} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) & = 0
\end{align}
%
The {\it direct} formulation\cite{YoonLenhoff1990} couples the two expressions of Eq. \eqref{eq:direct} on $\Gamma$, considering the interface conditions. Even though this formulation is poorly conditioned with respect to the mesh size, it can still model large systems when it is preconditioned appropriately.\cite{AltmanBardhanWhiteTidor09,MartinezETal2019,wang2021high} 
The formulation presented by Lu\cite{LuETal2006} takes the normal derivative of \eqref{eq:direct} to couple both $\phi$ and $\partial\phi/\partial\mathbf{n}$ as
%
\begin{align}\label{eq:lu}
    & \frac{\phi_{2,\Gamma}}{2}\left(\frac{\epsilon_1}{\epsilon_2}+1\right) - \left(K_Y^\Gamma - \frac{\epsilon_1}{\epsilon_2}K_L^\Gamma\right)(\phi_{2,\Gamma}) + \left(V_Y^\Gamma - V_L^\Gamma\right)\left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) = \sum_{k=0}^{N_q}  \frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}
     \nonumber \\
    &\frac{\epsilon_1}{\epsilon_2}\left(W_Y^\Gamma - W_L^\Gamma\right)(\phi_{2,\Gamma}) +  \frac{1}{2}\frac{\phi_{2,\Gamma}}{\partial\mathbf{n}}\left(1+\frac{\epsilon_1}{\epsilon_2}\right) + \left(\frac{\epsilon_1}{\epsilon_2}K_Y^{\prime\Gamma} - K_L^{\prime\Gamma}\right)\left( \frac{\partial}{\partial \mathbf{n}} \phi_{2,\Gamma} \right) = \sum_{k=0}^{N_q}  \frac{\partial}{\partial\mathbf{n}_\mathbf{r}}\left(\frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}\right)
\end{align}
%
The two new integral operators that appear are defined as 
%
\begin{align}\label{eq:adj_hyp}
K^{\prime\Gamma}_{L,Y}(\varphi) = \oint_\Gamma \frac{g_{L,Y}}{\partial\mathbf{n}}(\mathbf{r}_\Gamma,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
W^\Gamma_{L,Y}(\varphi) = - \oint_\Gamma \frac{\partial^2 g_{L,Y}}{\partial\mathbf{n}'\partial\mathbf{n}}(\mathbf{r}_\Gamma,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'
\end{align}
%
and are called the {\it  adjoint double layer} and {\it hypersingular} operators, respectively. Note that the potential in Eq. \eqref{eq:lu} is evaluated on the exterior.

Having computed the electrostatic potential with Eq. \eqref{eq:lu}, we obtain the reaction potential ($\phi_r$) in Eq. \eqref{eq:dG} by subtracting out the Coulombic component out of $\phi_1$ in Eq. \eqref{eq:volume_potential}. This gives
%
\begin{align} \label{eq:phi_reac}
\phi_{r} = - K_{L}^{\Omega_1}(\phi_{1,\Gamma}) + V_{L}^{\Omega_1} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right),
\end{align}
%
which is valid as long as $\epsilon_1$ is a constant.

>>>>

There are several possible boundary integral formulations of Equation \eqref{eq:pbe}.~\cite{search2022towards} The simplest form was presented by Yoon ºand Lenhoff in 1991,~\cite{YoonLenhoff1990} known as the \emph{direct} formulation, however, the resulting system is usually ill-conditioned. Here, we use the better conditioned formulation presented by Lu and co-workers.~\cite{LuETal2006,LuETal2009,debuhr2016dashmm}
Applying Green's second identity to Equation \eqref{eq:pbe} we get 
%
\begin{align} \label{eq:volume_potential}
\phi_{1}(\mathbf{r}_{\Omega_1})+ K_{L}^{\Omega_1}(\phi_{1,\Gamma}) -  V_{L}^{\Omega_1} \lambda_1 & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Omega_1} - \mathbf{r}_k|}  \quad \text{on $\Omega_1$,} \nonumber \\
\phi_{2}(\mathbf{r}_{\Omega_2}) - K_{Y}^{\Omega_2}(\phi_{2,\Gamma}) + V_{Y}^{\Omega_2} \lambda_2& = 0 \quad \text{on $\Omega_2$,}
\end{align}
%
where $\mathbf{r}_\Omega$ is evaluated anywhere in the correspoding domain, except $\Gamma$, and $\phi_{1,\Gamma} = \phi_1(\mathbf{r}_\Gamma)$ and $\phi_{2,\Gamma} = \phi_2(\mathbf{r}_\Gamma)$ are the potentials on $\Gamma$ as we approach from $\Omega_1$ and $\Omega_2$, respectively. For convenience, we call the normal derivative on the interface ({\it ie.} the Neumann trace) $\lambda=\left(\frac{\partial}{\partial \mathbf{n}}  \phi_{\Gamma}  \right)$. $K$ and $V$ are the double- and single-layer potentials for the Laplace (subscript $L$) and Yukawa (subscript $Y$, also known as modified Helmholtz) kernels
%
\begin{align}\label{eq:single_double}
V^\Omega_{L,Y}(\varphi) = \oint_\Gamma g_{L,Y}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
K^\Omega_{L,Y}(\varphi) = \oint_\Gamma \frac{\partial g_{L,Y}}{\partial\mathbf{n}'}(\mathbf{r}_\Omega,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}',\nonumber\\
\end{align}
%
where $\varphi(\mathbf{r})$ can be any distribution over $\Gamma$, and 
%
\begin{align}\label{eq:green_func}
g_L(\mathbf{r},\mathbf{r}')=\frac{1}{4\pi|\mathbf{r}-\mathbf{r}'|} \nonumber \\
g_Y(\mathbf{r},\mathbf{r}')=\frac{e^{-\kappa|\mathbf{r}-\mathbf{r}'|}}{4\pi|\mathbf{r}-\mathbf{r}'|}
\end{align}
%
are the free-space Green's function of the Laplace and linearized Poisson-Boltzmann equations, respectively.

If we take the limit of Eq. \eqref{eq:volume_potential} as $\mathbf{r}_{\Omega}$ approaches $\Gamma$, we get
%
\begin{align} \label{eq:direct}
\frac{\phi_{1,\Gamma}}{2}+ K_{L}^{\Gamma}(\phi_{1,\Gamma}) -  V_{L}^{\Gamma} \lambda_1 & = \frac{1}{\epsilon_1} \sum_{k=0}^{N_q}  \frac{q_k}{4\pi|\mathbf{r}_{\Gamma} - \mathbf{r}_k|} \nonumber \\
\frac{\phi_{2,\Gamma}}{2} - K_{Y}^{\Gamma}(\phi_{2,\Gamma}) + V_{Y}^{\Gamma} \lambda_2 & = 0
\end{align}
%
The {\it direct} formulation\cite{YoonLenhoff1990} couples the two expressions of Eq. \eqref{eq:direct} on $\Gamma$, considering the interface conditions. Even though this formulation is poorly conditioned with respect to the mesh size, it can still model large systems when it is preconditioned appropriately.\cite{AltmanBardhanWhiteTidor09,MartinezETal2019,wang2021high} 
The formulation presented by Lu\cite{LuETal2006} takes the normal derivative of \eqref{eq:direct} to couple both $\phi$ and $\partial\phi/\partial\mathbf{n}$ as
%
\begin{align}\label{eq:lu}
    & \frac{\phi_{2,\Gamma}}{2}\left(\frac{\epsilon_1}{\epsilon_2}+1\right) - \left(K_Y^\Gamma - \frac{\epsilon_1}{\epsilon_2}K_L^\Gamma\right)(\phi_{2,\Gamma}) + \left(V_Y^\Gamma - V_L^\Gamma\right)\lambda_2 = \sum_{k=0}^{N_q}  \frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}
     \nonumber \\
    &\frac{\epsilon_1}{\epsilon_2}\left(W_Y^\Gamma - W_L^\Gamma\right)(\phi_{2,\Gamma}) +  \frac{1}{2}\lambda_2\left(1+\frac{\epsilon_1}{\epsilon_2}\right) + \left(\frac{\epsilon_1}{\epsilon_2}K_Y^{\prime\Gamma} - K_L^{\prime\Gamma}\right)\lambda_2 = \sum_{k=0}^{N_q}  \frac{\partial}{\partial\mathbf{n}_\mathbf{r}}\left(\frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}\right)
\end{align}
%
The two new integral operators that appear are defined as 
%
\begin{align}\label{eq:adj_hyp}
K^{\prime\Gamma}_{L,Y}(\varphi) = \oint_\Gamma \frac{g_{L,Y}}{\partial\mathbf{n}}(\mathbf{r}_\Gamma,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'\nonumber\\
W^\Gamma_{L,Y}(\varphi) = - \oint_\Gamma \frac{\partial^2 g_{L,Y}}{\partial\mathbf{n}'\partial\mathbf{n}}(\mathbf{r}_\Gamma,\mathbf{r}')\varphi(\mathbf{r}')d\mathbf{r}'
\end{align}
%
and are called the {\it  adjoint double layer} and {\it hypersingular} operators, respectively. Note that the potential in Eq. \eqref{eq:lu} is evaluated on the exterior. 
In weak form, Eq. \eqref{eq:lu} becomes
%
\begin{align}\label{eq:lu_disc}
&\left< \frac{1}{2}\left(\frac{\epsilon_1}{\epsilon_2}+1\right) I - \left(K_Y^\Gamma - \frac{\epsilon_1}{\epsilon_2}K_L^\Gamma\right)(\phi_{2}), v \right>_{\Gamma} + \left< \left(V_Y^\Gamma - V_L^\Gamma\right) \lambda_2, v \right>_\Gamma = \left( \sum_{k=0}^{N_q}  \frac{q_k}{4\pi\epsilon_2|\mathbf{r}_{\Gamma} - \mathbf{r}_k|}, v \right)_{\Omega_1} \nonumber \\
&\left< \frac{\epsilon_1}{\epsilon_2}\left(W_Y^\Gamma - W_L^\Gamma\right)\phi_{2} \right>_\Gamma + \left< \left(1+\frac{\epsilon_1}{\epsilon_2}\right) + \left(\frac{\epsilon_1}{\epsilon_2}K_Y^{\prime\Gamma} - K_L^{\prime\Gamma}\right)\lambda_2, v\right>_\Gamma = 0
\end{align}
%
where $v$ is a test function, and $\left<\varphi,v\right>_\Gamma = \int_\Gamma \varphi(\mathbf{r})v(\mathbf{r})d\mathbf{r}$ and $\left(\varphi,v\right)_\Omega = \int_\Omega \varphi(\mathbf{r})v(\mathbf{r})d\mathbf{r}$ are the inner products on the surface and domain, respectively. This can be written in matrix form as
%
\begin{align}
Matrix equation
\end{align}
%

Having computed the electrostatic potential with Eq. \eqref{eq:lu_disc}, we obtain the reaction potential ($\phi_r$) in Eq. \eqref{eq:dG} by subtracting out the Coulombic component out of $\phi_1$ in Eq. \eqref{eq:volume_potential}. This gives
%
\begin{align} \label{eq:phi_reac}
\phi_{r} = - K_{L}^{\Omega_1}(\phi_{1,\Gamma}) + V_{L}^{\Omega_1} \left(\frac{\partial}{\partial \mathbf{n}}  \phi_{1,\Gamma}  \right),
\end{align}
%
which is valid as long as $\epsilon_1$ is a constant.

### 2022-03-29 results.tex
added more refs and weak form in bembem

In implicit-solvent models, the molecule is usually considered as a region with constant permittivity, in this case, with $\epsilon_1=4$.
In the solvent region, we used the permittivity of water ($\epsilon_2$=80) and an inverse of the Debye length of $\kappa=0.125$ \AA$^{-1}$.

>>>>

In implicit-solvent models, the molecule is usually considered as a region with constant permittivity, in this case, with $\epsilon_1=4$.
In the solvent region, we used the permittivity of water ($\epsilon_2$=80) and an inverse of the Debye length of $\kappa=0.125$ \AA$^{-1}$.
As in these cases there is an analytical solution for $\phi_c$ in Eq. \eqref{eq:phic}, we compute $\phi_r$ with Eq. \eqref{eq:phi_reac} in both BEM-BEM and FEM-BEM coupling approaches. Then, for FEM-BEM, the integral over $\Gamma$ in Eq. \eqref{eq:phi_reac} corresponds to the trace of the solution vector.
